### 2014-06-26 manuscript/manuscript.tex
Fixed refs.

The paucity of annotations is understandable, as RNA gene annotation
is non-trivial \cite{Freyhult:2007,Nawrocki:2009}. However, the
increasing roles for RNA uncovered through experimental and
bioinformatic study make illuminating this dark matter all the more
urgent. Among the remarkable discoveries made are, riboswitch-mediated
regulation \cite{Barrick:2007}, transcriptional termination by RNA
elements \cite{Gardner:2011,Chen:2013,Cambray:2013}, identification of
novel natural catalytic RNAs \cite{Roth:2014}, CRISPR-mediated
acquired immunity \cite{Barrangou:2007}, temperature-dependent gene
regulation \cite{Loh:2013}, and sno-like RNAs in Archaea
\cite{Omer:2000,Gardner:2010}. The Rfam database
\cite{Gardner:2011,Burge:2013} provides a valuable platform for
collating and characterising these and other families of noncoding
RNA. However, a recent comparative analysis \cite{Hoeppner:2012}
revealed that fewer than 7\% of RNA families within Bacteria and less
than 19\% in Archaea show a broad phylogenetic distribution (that is,
presence in at least 50\% of sequenced phyla). Crucially, that
analysis revealed that underlying genome sequencing biases were a
major contributor to this pattern, and that the wider genomic sampling
provided by the GEBA dataset did help improve identification of
broadly-conserved RNA families \cite{Hoeppner:2012}. Tools such as
RNA-seq \cite{Croucher:2010} and transposon insertion
sequencing \cite{van_Opijnen:2013,Barquist:2013,Barquist:2013a}
promise to complement comparative genomics tools for RNA family
discovery, and it may be possible to use a mix of data types in the
identification of RNA elements. However, to date, no systematic
analysis of available data has been undertaken, suggesting ncRNAs may
be hidden in the deluge of published data.

@@

To assess the relative contribution of noncoding RNAs and
protein-coding genes to transcriptional output, we collected all
publicly-available bacterial and archaeal RNA-seq datasets (available
as of {\bf August 2013}), spanning 37 species/strains and 400
datasets. For all datasets, we supplemented publicly available genome
annotations with screening for additional loci against Pfam and Rfam
databases, followed by manual identification of expressed unannotated
regions. This latter annotation yielded 922 expressed RUFs (RNAs of
unknown function).

@@

Effective comparative analysis requires appropriate phylogenetic
distances between species under investigation (Eddy 2005). The
transcriptomes available for our analyses are limited in number and
phylogenetic diversity, and efforts to redress phylogenetic biases in
genome sequencing have focused on maximising phylogenetic diversity
across the tree of life \cite{Wu:2009,Rinke:2013}. Moreover,
underlying biases in genome sampling have been shown to affect the
assessment of ncRNA conservation \cite{Hoeppner:2012}. We therefore
considered how species selection impacts comparative analysis as a
tool for distinguishing signals of conservation in ncRNAs from
transcriptional noise.

@@

{\bf Expression of core genes:} We defined a set of 40 core
protein-coding genes based on \cite{Wu:2013} and 16 noncoding RNA
genes (the union of tRNA, RNaseP, tmRNA, SRP, 6S and rRNA RNA
families) \cite{Gardner:2011,Burge:2013}. If the median read depth
greater than 10X, we defined the gene as expressed. For each dataset,
we report the fraction of the 40 core genes that are expressed.

@@

>>>>

The paucity of annotations is understandable, as RNA gene annotation
is non-trivial \cite{Freyhult:2007,Nawrocki:2009}. However, the
increasing roles for RNA uncovered through experimental and
bioinformatic study make illuminating this dark matter all the more
urgent. Among the remarkable discoveries made are: riboswitch-mediated
regulation \cite{Barrick:2007}, transcriptional termination by RNA
elements \cite{Gardner:2011,Chen:2013,Cambray:2013}, identification of
novel natural catalytic RNAs \cite{Roth:2014}, CRISPR-mediated
acquired immunity \cite{Barrangou:2007}, temperature-dependent gene
regulation \cite{Loh:2013}, and sno-like RNAs in Archaea
\cite{Omer:2000,Gardner:2010}. The Rfam database
\cite{Gardner:2011,Burge:2013} provides a valuable platform for
collating and characterising these and other families of noncoding
RNA. However, a recent comparative analysis \cite{Hoeppner:2012}
revealed that fewer than 7\% of RNA families within Bacteria and less
than 19\% in Archaea show a broad phylogenetic distribution (that is,
presence in at least 50\% of sequenced phyla). Crucially, that
analysis revealed that underlying genome sequencing biases were a
major contributor to this pattern, and that the wider genomic sampling
provided by the GEBA dataset did help improve identification of
broadly-conserved RNA families \cite{Hoeppner:2012}. Tools such as
RNA-seq \cite{Croucher:2010} and transposon insertion
sequencing \cite{van_Opijnen:2013,Barquist:2013,Barquist:2013a}
promise to complement comparative genomics tools for RNA family
discovery, and it may be possible to use a mix of data types in the
identification of RNA elements. However, to date, no systematic
analysis of available data has been undertaken, suggesting ncRNAs may
be hidden in the deluge of published data.

@@

To assess the relative contribution of noncoding RNAs and
protein-coding genes to transcriptional output, we collected all
publicly-available bacterial and archaeal RNA-seq datasets (available
as of {\bf August 2013}), spanning 37 species/strains and 400
datasets. For all datasets, we supplemented publicly available genome
annotations with screening for additional loci against Pfam and Rfam
databases, followed by manual identification of expressed unannotated
regions. This latter annotation yielded 922 expressed RNAs of
Unknown Function (RUFs).

@@

Effective comparative analysis requires appropriate phylogenetic
distances between species under investigation \cite{Eddy:2005}. The
transcriptomes available for our analyses are limited in number and
phylogenetic diversity, and efforts to redress phylogenetic biases in
genome sequencing have focused on maximising phylogenetic diversity
across the tree of life \cite{Wu:2009,Rinke:2013}. Moreover,
underlying biases in genome sampling have been shown to affect the
assessment of ncRNA conservation \cite{Hoeppner:2012}. We therefore
considered how species selection impacts comparative analysis as a
tool for distinguishing signals of conservation in ncRNAs from
transcriptional noise.

@@

{\bf Expression of core genes:} We defined a set of 40 core
protein-coding genes based on \cite{Wu:2013,Darling:2014} and 16 noncoding RNA
genes (the union of tRNA, RNaseP, tmRNA, SRP, 6S and rRNA RNA
families) \cite{Gardner:2011,Burge:2013}. If the median read depth
greater than 10X, we defined the gene as expressed. For each dataset,
we report the fraction of the 40 core genes that are expressed.

@@

\newpage
\clearpage
